Towards a Neural Co-Processor Which Restores Movement After Stroke: Modeling a Proof-of-Concept

### 2021-11-14 paper/article.tex
Paper progress

While open loop stimulation techniques have yielded clinically useful results,
their results in many domains have been mixed, such as use in visual
prostheses \cite{bosking.visual}, and use in invoking somatosensory feedback
\cite{cronin.tact}. Likely this is due to the stimulation not being conditioned
on the ongoing dynamics of the circuit being stimulated. Moment-to-moment and
throughout the day, the circuit will respond differently to the same stimulus,
as a result of differing inputs and ongoing activity. Stimulation
therefore needs to be proactively adapted in response. This need is even
greater over longer time scales as the effects of plasticity and ageing change
the connectivity of the brain.

@@

Note that the EN effectively constitutes a stimulation model. Its design is
somewhat motivated by the common linear time-invariant state space model of
stimulation, as in e.g. Yang et al. \cite{shanechi.stimmodel}, which is a helpfully
simple (linear) approach. That approach seeks to model neural network dynamics in
terms of a stationary linear model, which, due in part to its simplicity and the
number of techniques developed for it, lends itself to useful interpretations and
analyses of those neural dynamics.

There are some key differences between that approach and the EN architecture we
present here. Our EN's purpose is simply to train the CPN, so we pick an
architecture we find to be best suited to that job. First, our EN architecture
is not linear, through the use of a $tanh$ activation function. That allows us
to escape some of the limits in representational power of linear models, though
we didn't deeply test a strictly linear approach.
Second, LSTM cells are designed to better capture the relevant asepcts of
long-running network dynamics, and in our case we found that they were
crucial for that reason: when using vanilla RNN cells, our co-processor
wasn't able to train well. Compare that to the simpler cells commonly
used in vanilla RNNs, which yield functionally the same approach as the
linear state space model, less the use of a nonlinear activation function.

@@

\subsection{Training Algorithm}
In all experiments, training the co-processor requires a careful interleaving of
EN and CPN training epochs. EN training epochs concentrate on updating the EN,
based on observations of the effect of stimulation on the simulated network's output.
The loss function when training the EN is the MSE loss between the actual and
predicted output of the mRNN (i.e. muscle velocities). We then use the EN
to train the CPN during the CPN training epochs. During that time, we backpropagate
the MSE loss between the EN's predicted output, and target output, to the CPN. In
other words: we treat the EN output as a proxy to the true brain, allowing us
to train the CPN, since we have no way to ``backpropagate through the brain'' itself.
See Fig. \ref{fig:training}.

@@

\subsubsection{EN training}
For the EN to be useful, it must accurately predict the effect of stimulation
produced by the CPN. In addition to that, backpropagating through it must yield
gradients which train the CPN to output more useful stimulus. We discovered that
this latter property does not necessarily occur simply by virtue of the former.
We have found that an EN can be trained to extreme levels of predictive power, even
on random stimulation, and to orders of magnitude lower loss than the task loss,
while at the same time backpropagation through it yields gradients which are not
useful for training the CPN. As a result, if the EN is not trained in specific ways,
the CPN exhibits unstable training. We have thus far been unable to discover a way
to measure and optimize for this second property, though pursuing it may be a
fruitful subject of future applied ML research. Instead of attempting to train an EN
directly to exhibit this property, we discovered a training regime that reliably makes
it occur.

The central concept of our training regime is to provide a carefully chosen variety of
training examples, and to use heavy weight regularization, in order to force
the EN to generalize well to stimulation which it hasn't yet seen. We hypothesize that
the crux of our EN training problem is one of over-fitting: that the EN may be trained
to great predictive power on a set of stimulation examples, but that the effective
dimensionality of the problem it is learning is comparatively quite high, due to both
the dimensionality of our stimulation parameters $\theta$, and the dynamics of the
network being stimulated. As a result, the hypersurface which it has fit to the training
examples may exhibit gradients not reflective of reality.

First, and most significantly, we structure the training data set for each EN training epoch
in a very specific way. Each epoch, we include:
\begin{itemize}
	\item examples using stimulation from the current CPN
	\item a small (100s) collection of CPNs which are copies of the current CPN, with
	      Gaussian-distributed mean-0 noise added to their parameters
	\item examples of white noise stimulation
\end{itemize}
We developed this structure under the belief that we needed to cover a sufficient
variety of examples to prevent overfit, and to do so in a way that emphasizes the neighborhood
in CPN parameter space of the current CPN. We initially attempted to pad our dataset with white
noise examples alone, but that was not sufficient to stabilize CPN training, despite the EN's
ability to reach a high predictive power. Adding in the addition of samples from the parameter
space near the current CPN appears to help ensure that the EN is localized enough to give
useful gradients for that area. As we will explain below, this produces an additional problem
that the EN is too specialized to the local area of CPN parameter space, but we then solve
that problem using EN retraining, interleaved with the CPN training.

@@

Second, we use PyTorch's $AdamW$ optimizer, which includes an option for weight regularization.
Additionally, we use a carefully chosen learning rate schedule. Our intuition is to leverage regularization as
one of the standard mitigations for overfit. To converge, the learning rate schedule also needs
to be chosen carefully: a relatively high learning rate at first, dropping to a very low rate soon
thereafter. Using too-low or too-high of a learning rate in any phase causes the EN learning
to not converge. EN training proceeds until it reaches a threshold of prediction error, defined as
a fraction of the current CPN's task loss.

\subsubsection{CPN training}
Once we discovered an effective way to train an EN, CPN training was straightforward. For this phase,
we generate training examples using the current CPN alone. The EN generates predictions of the
effect of those stimulations, and we backpropagate through the EN to generate training gradients
for the CPN. As with the EN, we constitute the dataset with a large random sample of trials from
the original Michaels task. See Fig. \ref{fig:training}.

@@

\subsubsection{Interleaved CPN/EN training, and adapting to non-stationarity}
Having defined the training epochs for the EN and CPN, we can define a training
algorithm on the basis of those two. We train the two in alternation, creating
a new EN each time we enter an EN training period. We explored the possibility
of reusing an existing EN by retraining it, for the purpose of training efficiency.
However, we found that retraining an EN was no more fast than training a new one,
and in-fact usually took longer, despite significant experimentation with learning
rates.

The key challenge we face is determining when an EN is no longer suitable for training the
current CPN. Expiring an EN at the right time accomplishes two things. First, it ensures
that our current EN is trained properly for the current CPN; without that, the CPN training
will become unstable. Second, expiring the EN is how we adapt our learning to the brain's
non-stationarity. At some point, the brain will have changed sufficiently for our EN to be
out-of-date, and therefore requiring replacement.

TODO: talk about retraining more

@@

Unfortunately, because we expire the EN rather than retraining it, and because we need
to expire it so often, training is less efficient than it otherwise could be. We explore
this some in Section \ref{sec:results}. In the next subsection, we illustrate one way we
mitigate training inefficiency.

TODO: pseudocode

\subsubsection{Optimization: data reuse}
One obvious

@@

Using the simulation described above, we perform three experiments to
demonstrate the co-processor's ability to learn:
\begin{itemize}
	\item 50\% AIP loss
	\item 100\% connectivity lesion between the F5 and M1 modules
	\item 100\% connectivity lesion between the F5 and M1 modules without brain co-adpatation
	\item 50\% M1 loss
\end{itemize}

Additionally, we perform each of the ...

TODO: train vs test

Summary of tunables:
\begin{itemize}
	\item Learning rate schedules
	\item Activation functions
	\item Obs function dimensionality
	\item Stim dimensionality
	\item Dimensionality of EN, CPN
\end{itemize}

@@

\begin{itemize}
	\item Task performance improves drastically.
		\begin{itemize}
			\item Fig: normalized task loss vs CPN training epoch
			\item Fig: normalized task loss vs training epoch
			\item Fig: normalized task loss vs training epoch
			\item Split by experiment
		\end{itemize}
	\item Fine tuning takes a long time
	\item Object classes separate. Fig: sep vs experiment vs epoch
	\item Can co-adapt with the brain, as it changes
	\item Training efficiency analysis. Show num epochs with and without recycling.
\end{itemize}

\section{Discussion and Conclusion}
\begin{itemize}
	\item Training efficiency
	\item Spectrum from simple low dimensional stimulation vectors today to higher dimensional future
	\item Toward in vivo application: example of lower dimensional problems this may help with
\end{itemize}

\section{Acknowledgements}
Ganguly, Priya, Anca, Justin, Luciano

\section{Ethical Statement}
asdf

>>>>

While open loop stimulation techniques have yielded clinically useful results,
their results in many domains have been mixed, such as use in visual
prostheses \cite{bosking.visual}, and use in invoking somatosensory feedback
\cite{cronin.tact}. Likely this is due to the stimulation not being conditioned
on the ongoing dynamics of the circuit being stimulated. Moment-to-moment and
throughout the day, the circuit will respond differently to the same stimulus,
as a result of differing inputs and ongoing activity. Stimulation
therefore needs to be proactively adapted in response. This need is even
greater over longer time scales as the effects of plasticity and ageing change
the connectivity of the brain. Closed-loop stimulation may also provide means
to better regulate the energy use of an implanted stimulator, allowing it to
intelligently regulate when to apply stimulation, and to balance implant
battery life and stimulation effect. Closed-loop stimulation also offers an
opportunity minimize side-effects of stimulation, through real time regulation
of the stimulation parameters, such as in the use of deep brain stimulation in
Parkinson's patients.

@@

Note that the EN effectively constitutes a stimulation model. We initially attempted
an EN architecture resembling the common linear time-invariant state space model of
stimulation, as in e.g. Yang et al. \cite{shanechi.stimmodel}, which is a helpfully
simple (linear) approach. That approach seeks to model neural network dynamics in
terms of a stationary linear model, which, due in part to its simplicity and the
number of techniques developed for it, lends itself to useful interpretations and
analyses of those neural dynamics. We found, however, that the linear approach,
as well as the related vanilla RNN approach with nonlinearities added, was not
sufficiently expressive of long-term dependencies, such that the co-processor
was capable of learning well.

Separately, nonlinear LSTM models have also been applied to modeling natural
neural networks, such as use in predicting local field potentials
\cite{kim.lstm}, and for stimulation modeling \cite{guclu.lstm}.
LSTM cells are designed to better capture the relevant aspects of
long-running network dynamics, and in our case we found that they were crucial
for that reason. As a result, both our EN and CPN models leverage LSTM cells.

@@

\subsection{Training Algorithm}
In all experiments, training the co-processor requires a careful interleaving of
EN and CPN training epochs. EN training epochs concentrate on updating the EN,
based on observations of the effect of stimulation on the simulated network's output.
The loss function when training the EN is the mean squared error (MSE) loss between the
actual and predicted output of the mRNN (i.e. muscle velocities). We then use the EN
to train the CPN during the CPN training epochs. During that time, we backpropagate
the MSE loss between the EN's predicted output, and target output, to the CPN. In
other words: we treat the EN output as a proxy to the true brain, allowing us
to train the CPN, since we have no way to ``backpropagate through the brain'' itself.
See Fig. \ref{fig:training}.

@@

\subsubsection{EN training}
For the EN to be useful, it must accurately predict the effect of stimulation
produced by the CPN. In addition to that, backpropagating through it must yield
gradients which train the CPN to output more useful stimulation. We discovered that
this latter property does not necessarily occur simply by virtue of the former.
We have found that an EN can be trained to extreme levels of predictive power, even
on random stimulation, and to orders of magnitude lower loss than the task loss,
while at the same time backpropagation through it yields gradients which are not
useful for training the CPN. As a result, if the EN is not trained in specific ways,
the CPN exhibits unstable training. We have thus far been unable to discover a way
to measure and optimize for this second property directly, though pursuing it may be a
fruitful subject of future applied ML research. Instead of attempting to train an EN
directly to exhibit this property, we discovered a training regime that indirectly,
but reliably, makes it occur.

The central concept of our training regime is to provide a carefully chosen variety of
training examples, and to use heavy weight regularization, in order to force
the EN to generalize well to stimulation which it hasn't yet seen. We hypothesize that
the crux of our EN training problem is one of over-fitting: that the EN may be trained
to high predictive power on a set of stimulation examples, but that the effective
dimensionality of the problem it is learning is comparatively quite high, due to both
the dimensionality of our stimulation parameters $\theta$, and the dynamics of the
network being stimulated. As a result, the hypersurface which it fits to the training
examples may exhibit gradients not reflective of reality.

First, and most significantly, we structure the training dataset for each EN training epoch
in a very specific way. Each epoch, we include:
\begin{itemize}
	\item examples using stimulation from the current CPN
	\item examples from a small (100s) collection of CPNs which are copies of the
	      current CPN, with Gaussian-distributed mean-0 noise added to their parameters
	\item examples of white noise stimulation
\end{itemize}
We developed this structure under the belief that we needed to cover a sufficient
variety of examples to prevent overfit, and to do so in a way that emphasizes the neighborhood
in CPN parameter space of the current CPN. We initially attempted to pad our dataset with white
noise examples alone, but that was not sufficient to stabilize CPN training, despite the EN's
ability to reach a high predictive power. Adding in the addition of samples from the parameter
space near the current CPN appears to help ensure that the EN is localized enough to give
useful gradients for that area. As we will explain below, this produces an additional problem
that the EN is too specialized to the local area of CPN parameter space, but we then solve
that problem using EN retraining, interleaved with the CPN training.

@@

Addiontally, we use PyTorch's $AdamW$ optimizer, which includes an option for weight regularization.
With that, we use a carefully chosen learning rate schedule. Our intuition is to leverage regularization as
one of the standard mitigations for overfit. To converge, the learning rate schedule needs
to be chosen carefully: a relatively high learning rate at first, dropping to a very low rate soon
thereafter. Using too-low or too-high of a learning rate in any phase causes the EN learning
to not converge. EN training proceeds until it reaches a threshold of prediction error, defined as
a fraction of the current CPN's task loss.

\subsubsection{CPN training}
Once an EN is properly trainedx, CPN training is straightforward. For this phase,
we generate training examples using the current CPN alone. The EN generates predictions of the
effect of those stimulations, and we backpropagate through the EN to generate training gradients
for the CPN. As with the EN, we constitute the dataset with a large random sample of trials from
the original Michaels task. See Fig. \ref{fig:training}.

@@

\subsubsection{Interleaved CPN/EN training, and adapting to non-stationarity}
Having defined the training epochs for the EN and CPN, we can define a training
algorithm on the basis of those two. We train the two in alternation, creating
a new EN each time we enter an EN training period. We explored the possibility
of reusing an existing EN by retraining it, for the purpose of training efficiency.
However, we found that retraining an EN was no more fast than training a new one,
and in-fact usually took longer, despite significant experimentation with learning
rates. It is unclear if that fact is an artifact of our simulation, or of the
problem more generally, and so it needs more examination in the future, for the
purpose of training efficiency.

The key remaining challenge is determining when an EN is no longer suitable for training the
current CPN. Expiring an EN at the right time accomplishes two things. First, it ensures
that our current EN is trained properly for the current CPN; without that, the CPN training
will become unstable. Second, expiring the EN is how we adapt our learning to the brain's
non-stationarity. At some point, the brain will have changed sufficiently for our EN to be
out-of-date, and therefore requiring replacement.

@@

\subsubsection{Improving learning efficiency with data reuse}
TODO. May drop this section if I can't get learning stabilized with this approach.

Using the simulation described above, we perform several experiments to
demonstrate the co-processor's ability to learn. In each experiment, we apply one
of the three lesion types described above. We additionally vary whether we simulate
brain co-adaptation, in order to better understand what the co-processor is learning.
Finally, we perform one experiment where we attempt to simulate natural recovery from
the lesion, prior to training the co-processor. In sum, our experiments are:

\begin{table}[]
\begin{tabular}{|l|l|c|c|}
\hline
\textbf{Index} & \textbf{Lesion}                                           & \multicolumn{1}{l|}{\textbf{\begin{tabular}[c]{@{}l@{}}Simulate\\ co-adaptation?\end{tabular}}} & \multicolumn{1}{l|}{\textbf{\begin{tabular}[c]{@{}l@{}}Simulate\\ pre-recovery?\end{tabular}}} \\ \hline
1              & 50\% AIP loss                                             &                                                                                                 &                                                                                                \\ \hline
2              & 50\% AIP loss                                             & X                                                                                               &                                                                                                \\ \hline
3              & 50\% M1 loss                                              &                                                                                                 &                                                                                                \\ \hline
4              & 50\% M1 loss                                              & X                                                                                               &                                                                                                \\ \hline
5              & 100\% loss of connectivity F5\textless{}-\textgreater{}M1 &                                                                                                 &                                                                                                \\ \hline
6              & 100\% loss of connectivity F5\textless{}-\textgreater{}M1 & X                                                                                               &                                                                                                \\ \hline
7              & 100\% loss of connectivity F5\textless{}-\textgreater{}M1 & X                                                                                               & X                                                                                              \\ \hline
\end{tabular}
\end{table}

In each experiment, we use a single mRNN model, sourced directly from the results
cited in Michaels \cite{michaels.mrnn}. We drive the experiment using the same set of
input and output data that the mRNN was trained on. The dataset contains a total of
502 trials, spread uniformally among 42 object classes. In each experiment, we separate
a random sample of 20\% of the dataset for validation.

@@

@@

TODO: summary of conclusions that will go here:
\begin{itemize}
	\item Task performance improves drastically. In all cases, performance is still improving slowly when I stop the simulation.
		\begin{itemize}
			\item 90\% on connection lesion, both with and without co-adaptation
			\item M1 50\%: 81\% recovery on first attempt with co-adaptation
			\item AIP 50\%: 35\% recovery w/co-adaptation. Presumably, this is because we've lost encoding functionality, and co-adaptation is very slow. If we ran the simulation for 1 month, who knows?
			\item Waiting on all other results.
		\end{itemize}
	\item Initially, performance improves orders of magnitude more quickly than later. 50-75\% recovery during this phase. This appears to be the co-processor learning the task structure (e.g. not to move the hand until after the hold signal is lifted).
	\item Fine tuning takes a far longer time
	\item This may have some connection to object class separation; show class sep metric side-by-side. (I have the data in hand, but haven't tried to correlate it to recovery yet).
	\item Can co-adapt with the brain, as it changes
	\item Training efficiency analysis. The majority of epochs (80\%; depends on experiment type) are EN training epochs, so most efficiency gains will be there.
	\item Compare co-adaptive case vs not. Some tasks will be impossible to accomplish without the two cooperating, if enough cortex has been lost. (waiting for new results to confirm)
	\item Show pre-recovery results; co-processor will improve beyond what recovery can achieve. (will run results this week)
\end{itemize}

\section{Discussion and Conclusion}
\begin{itemize}
	\item Co-processors can be trained, and they can adapt to a difficult problem
	\item One central issue that still needs work is data/training efficiency
	\item Pitch a couple ideas to help solve it:
		\begin{itemize}
			\item Continuous retraining of EN, rather than throwing it away
			\item Retain recent data and reuse it during training
		\end{itemize}
	\item Toward in vivo application: example of lower dimensional problems this may help with
\end{itemize}

\subsection{Learning efficiency}
Learning efficiency is a key concern for our design. Due to limits on patient fatigue,
limited time, implant battery life, and other concerns, it is not plausible to expect
a learning algorithm to have access to unlimited amounts of training data. As a result,
it is necessary to make efficient use of the data we can acquire.

One clear advantage of our co-processor approach is that, as in Bernal \cite{bernal.sim},
we can train our stimulation agent to a great degree

\subsection{Future application}
(on a lower dimensional problem presumably)

\section{Acknowledgements}
TODO: funding sources

The author would like to thank the following for discussions and insights related to
the present topic:
\begin{itemize}
	\item Karunesh Ganguly, UC San Francisco
	\item Priya Khanna, UC San Francisco
	\item Anca Dragan, UC Berkeley
	\item Justin Ong, University of Washington
	\item Luciano De La Iglesia, University of Washington
\end{itemize}

\section{Ethical Statement}
TODO.
